---
title: "3. Non-negative Tensor Factorization (`NTF` and `NTD`)"
author:
- name: Koki Tsuyuzaki
  affiliation: Laboratory for Bioinformatics Research,
    RIKEN Center for Biosystems Dynamics Research
  email: k.t.the-answer@hotmail.co.jp
- name: Itoshi Nikaido
  affiliation: Laboratory for Bioinformatics Research,
    RIKEN Center for Biosystems Dynamics Research
date: "`r Sys.Date()`"
bibliography: bibliography.bib
package: nnTensor
output: rmarkdown::html_vignette
vignette: |
  %\VignetteIndexEntry{3. Non-negative Tensor Factorization (`NTF` and `NTD`)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

In this vignette we consider approximating a non-negative tensor as a product of multiple non-negative low-rank matrices (a.k.a., factor matrices) and a core tensor.

Test data available from `toyModel`. Here, we set two datasets to simulate different situations.

```{r data, echo=TRUE}
library("nnTensor")
X_CP <- toyModel("CP")
X_Tucker <- toyModel("Tucker")
```

In the first case, you will see that there are four small blocks in the diagonal direction of the data tensor.

```{r plot_data, echo=TRUE, fig.height=4, fig.width=4}
plotTensor3D(X_CP)
```

In the second case, you will see that there are six long blocks in the data tensor.

```{r plot_data2, echo=TRUE, fig.height=4, fig.width=4}
plotTensor3D(X_Tucker)
```

# NTF

To decompose a non-negative tensor ($\mathcal{X}$), non-negative CP decomposition (a.k.a. non-negative tensor factorization; NTF [@ntf1; @ntf2; @ntf3]) can be applied. NTF appoximates $\mathcal{X}$ ($N \times M \times L$) as the mode-product of a core tensor $S$ ($J \times J \times J$) and factor matrices $A_1$ ($J \times N$), $A_2$ ($J \times M$), and $A_3$ ($J \times L$).

$$
\mathcal{X} \approx \mathcal{S} \times_{1} A_1 \times_{2} A_2 \times_{3} A_3\ \mathrm{s.t.}\ \mathcal{S} \geq 0, A_{k} \geq 0\ (k=1 \ldots 3)
$$

Note that \times_{k} is the mode-$k$ product [@amari] and the core tensor $S$ has non-negative values only in the diagonal element.

NTF can be performed as follows:

```{r ntf_cp, echo=TRUE}
set.seed(123456)
out_NTF_CP <- NTF(X_CP, algorithm="KL", rank=4)
str(out_NTF_CP, 2)
```
As same as other matrix factorization methods, the values of reconstruction error and relative error indicate that the optimization is converged.

```{r conv_ntf_cp, echo=TRUE, fig.height=4, fig.width=8}
layout(t(1:2))
plot(log10(out_NTF_CP$RecError[2:101]), type="b", main="Reconstruction Error")
plot(log10(out_NTF_CP$RelChange[2:101]), type="b", main="Relative Change")
```

By using `recTensor`, user can easily reconstruct the data from core tensor and factor matrices as follows.

```{r rec_ntf_cp, echo=TRUE, fig.height=4, fig.width=8}
recX_CP <- recTensor(out_NTF_CP$S, out_NTF_CP$A, idx=1:3)
layout(t(1:2))
plotTensor3D(X_CP)
plotTensor3D(recX_CP)
```

When applying `NTF` against the first data, we can see that the four blocks are properly reconstrcted. Next, we apply `NTF` aginst the second data.

```{r ntf_tucker, echo=TRUE}
set.seed(123456)
out_NTF_Tucker <- NTF(X_Tucker, algorithm="KL", rank=4)
str(out_NTF_Tucker, 2)
```

```{r conv_ntf_tucker, echo=TRUE, fig.height=4, fig.width=8}
layout(t(1:2))
plot(log10(out_NTF_Tucker$RecError[2:101]), type="b", main="Reconstruction Error")
plot(log10(out_NTF_Tucker$RelChange[2:101]), type="b", main="Relative Change")
```

Unlike with the first data, the second data shows that `NTF` does not work.

```{r rec_ntf_tucker, echo=TRUE, fig.height=4, fig.width=8}
recX_Tucker <- recTensor(out_NTF_Tucker$S, out_NTF_Tucker$A, idx=1:3)
layout(t(1:2))
plotTensor3D(X_Tucker)
plotTensor3D(recX_Tucker)
```

# NTD

Next, we introduce another type of non-negative tensor decomposition method, non-negative Tucker decomposition (NTD [@ntd1; @ntd2; @ntd3; @ntd4]). The difference with the NTF is that different ranks can be specified for factor matrices such as $A_1$ ($J1 \times N$), $A_2$ ($J2 \times M$), and $A_3$ ($J3 \times L$) and that the core tensor can have non-negative values in the non-diagonal elements. This degree of freedom will show that NTD fits the second set of data well.

```{r ntd, echo=TRUE}
set.seed(123456)
out_NTD <- NTD(X_Tucker, rank=c(3,4,5))
str(out_NTD, 2)
```

```{r conv_ntd, echo=TRUE, fig.height=4, fig.width=8}
layout(t(1:2))
plot(out_NTD$RecError[2:101], type="b", main="Reconstruction Error")
plot(out_NTD$RelChange[2:101], type="b", main="Relative Change")
```

```{r rec_ntd_tucker, echo=TRUE, fig.height=4, fig.width=8}
recX_Tucker2 <- recTensor(out_NTD$S, out_NTD$A, idx=1:3)
layout(t(1:2))
plotTensor3D(X_Tucker)
plotTensor3D(recX_Tucker2)
```

# NTD-2

NTD-2 extracts two factor matrices from two modes of a non-negative third-order tensor. By specifying two values against `rank` and `modes`, NTD-2 can be easily performed as follows. For mode not specified, a unit matrix is assigned instead.

```{r ntd2, echo=TRUE}
set.seed(123456)
out_NTD2_1 <- NTD(X_Tucker, rank=c(3,4), modes=1:2)
out_NTD2_2 <- NTD(X_Tucker, rank=c(3,5), modes=c(1,3))
out_NTD2_3 <- NTD(X_Tucker, rank=c(4,5), modes=2:3)
```

```{r conv_ntd2, echo=TRUE, fig.height=12, fig.width=8}
layout(rbind(1:2, 3:4, 5:6))
plot(out_NTD2_1$RecError[2:101], type="b", main="Reconstruction Error\nNTD-2 (mode=1:2)")
plot(out_NTD2_1$RelChange[2:101], type="b", main="Relative Change\nNTD-2 (mode=1:2)")
plot(out_NTD2_2$RecError[2:101], type="b", main="Reconstruction Error\nNTD-2 (mode=c(1,3))")
plot(out_NTD2_2$RelChange[2:101], type="b", main="Relative Change\nNTD-2 (mode=c(1,3))")
plot(out_NTD2_3$RecError[2:101], type="b", main="Reconstruction Error\nNTD-2 (mode=2:3)")
plot(out_NTD2_3$RelChange[2:101], type="b", main="Relative Change\nNTD-2 (mode=2:3)")
```

```{r rec_ntd_tucker2, echo=TRUE, fig.height=8, fig.width=8}
recX_Tucker2_1 <- recTensor(out_NTD2_1$S, out_NTD2_1$A, idx=1:3)
recX_Tucker2_2 <- recTensor(out_NTD2_2$S, out_NTD2_2$A, idx=1:3)
recX_Tucker2_3 <- recTensor(out_NTD2_3$S, out_NTD2_3$A, idx=1:3)
layout(rbind(1:2, 3:4))
plotTensor3D(X_Tucker)
plotTensor3D(recX_Tucker2_1)
plotTensor3D(recX_Tucker2_2)
plotTensor3D(recX_Tucker2_3)
```

# NTD-1

NTD-1 extracts a factor matrix from only one mode of a non-negative third-order tensor. By specifying only one value against `rank` and `modes`, NTD-1 can be easily performed as follows. For modes not specified, two unit matrices is assigned instead.

```{r ntd1, echo=TRUE}
set.seed(123456)
out_NTD1_1 <- NTD(X_Tucker, rank=3, modes=1)
out_NTD1_2 <- NTD(X_Tucker, rank=4, modes=2)
out_NTD1_3 <- NTD(X_Tucker, rank=5, modes=3)
```

```{r conv_ntd1, echo=TRUE, fig.height=12, fig.width=8}
layout(rbind(1:2, 3:4, 5:6))
plot(out_NTD1_1$RecError[2:101], type="b", main="Reconstruction Error\nNTD-1 (mode=1:2)")
plot(out_NTD1_1$RelChange[2:101], type="b", main="Relative Change\nNTD-1 (mode=1:2)")
plot(out_NTD1_2$RecError[2:101], type="b", main="Reconstruction Error\nNTD-1 (mode=c(1,3))")
plot(out_NTD1_2$RelChange[2:101], type="b", main="Relative Change\nNTD-1 (mode=c(1,3))")
plot(out_NTD1_3$RecError[2:101], type="b", main="Reconstruction Error\nNTD-1 (mode=2:3)")
plot(out_NTD1_3$RelChange[2:101], type="b", main="Relative Change\nNTD-1 (mode=2:3)")
```

```{r rec_ntd_tucker1, echo=TRUE, fig.height=8, fig.width=8}
recX_Tucker2_1 <- recTensor(out_NTD1_1$S, out_NTD1_1$A, idx=1:3)
recX_Tucker2_2 <- recTensor(out_NTD1_2$S, out_NTD1_2$A, idx=1:3)
recX_Tucker2_3 <- recTensor(out_NTD1_3$S, out_NTD1_3$A, idx=1:3)
layout(rbind(1:2, 3:4))
plotTensor3D(X_Tucker)
plotTensor3D(recX_Tucker2_1)
plotTensor3D(recX_Tucker2_2)
plotTensor3D(recX_Tucker2_3)
```

# Higher-order tensor decomposition

`NTF` and `NTD` can also be applied to non-negative higher-order tensors like follows.

```{r data2, echo=TRUE}
library("nnTensor")
library("rTensor")
X_Higher <- as.tensor(array(runif(3*4*5*6*7), dim=c(3,4,5,6,7)))
out_NTF_Higher <- NTF(X_Higher, rank=4)
out_NTD_Higher <- NTD(X_Higher, rank=c(1,2,1,2,3))
```

# Session Information {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

# References
